
\chapter{Flowers}


\verseref{44}
Who is it\\
that can truly see as they are\\
this earth, this body,\\
the hell realms and the heavenly realms?\\
Who can discern\\
the well-taught Path of Wisdom,\\
in the way the eye of a skilled florist\\
can select perfect blooms?

\verseref{45}
It is those who knowingly walk the Way\\
who can see as they truly are\\
this earth, this body,\\
the hell realms and the heavenly realms.\\
They are the ones who can discern\\
the well-taught Path of Wisdom.


\verseref{46}
Know the body\\
to be as transient as foam, a mirage.\\
The flower of sensual passion\\
has a hidden barb.\\
See this and pass beyond death.


\verseref{47}
As a flash flood\\
can sweep away a sleeping village,\\
so death can destroy\\
those who only seek the flowers\\
of casual sensual pleasures.


\verseref{48}
To be frustrated and ungratified,\\
yet still only seek the flowers\\
of casual sensual pleasures\\
brings one under the destroyer's sway.


\verseref{49}
As a bee gathering nectar\\
does not harm or disturb\\
the colour and fragrance of the flower,\\
so do the wise move\\
through the world.


\verseref{50}
Dwell not on the faults\\
and shortcomings of others,\\
instead, seek clarity\\
about your own.


\verseref{51}
As a beautiful flower\\
without fragrance is disappointing,\\
so are wise words\\
without right action.



\verseref{52}
As a beautiful flower\\
with a delightful fragrance is pleasing,\\
so is wise and lovely speech\\
when matched with right action.


\verseref{53}
As many garlands can be made\\
from a heap of flowers,\\
so too, much that is wholesome can be done\\
during this human existence.


\verseref{54}
The fragrance of flowers or sandalwood\\
blows only with the prevailing wind,\\
but the fragrance of virtue\\
pervades all directions.


\verseref{55}
The fragrance of virtue\\
surpasses by far\\
the fragrance of flowers\\
or sandalwood.


\verseref{56}
The aroma of sandalwood\\
and the scent of flowers\\
bring only little delight\\
compared to the fragrance of virtue,\\
which suffuses even the heavenly realms.


\verseref{57}
It is not possible for Mara to find\\
those who abide in appreciative awareness,\\
who by perfect knowledge are freed\\
and live in virtue.

\verseref{58-59}
Just as a sweet-smelling and beautiful lotus\\
can grow from a pile of discarded waste,\\
the radiance of a true disciple of the Buddha\\
outshines dark shadows cast by ignorance.

