
\chapter{Awareness}

\verseref{21}
Appreciative awareness leads to life;\\
heedless avoidance is the path to death.\\
Those who are aware are fully alive,\\
while those who are heedless\\
are as if already dead.

\verseref{22}
The wise, being fully alive, rejoice\\
in appreciative awareness,\\
and abide delighting\\
in this capacity.


\verseref{23}
The Awakened Ones, firm in their resolve,\\
vigorously apply themselves,\\
and know freedom from bondage:\\
liberation, true security.


\verseref{24}
Those who are energetically\\
committed to the Way,\\
who are pure and considerate in effort,\\
composed and virtuous in conduct,\\
steadily increase in radiance.


\verseref{25}
By endeavour, vigilance,\\
restraint and self-control,\\
let the wise make islands of themselves\\
which no flood can overwhelm.



\verseref{26}
Those who are foolish and confused\\
betray themselves to heedlessness.\\
The wise treasure the awareness\\
they have cultivated\\
as their most precious possession.


\verseref{27}
"Do not become lost in negligence,\\
do not become lost in sensuality."\\
Heeding such council,\\
the contemplative discovers\\
profound release.

\verseref{28}
Those Awakened Ones,\\
who have tasted freedom\\
from all distraction\\
by cultivating awareness\\
view all who are suffering\\
with compassionate perspective,\\
as one on a mountaintop\\
is able to view the plains.


\verseref{29}
Aware among those who are heedless,\\
awake among those who sleep,\\
the wise go forward like strong young horses,\\
leaving the exhausted behind.


\verseref{30}
By way of diligent awareness\\
the god Magha won his realm.\\
Diligence is ever rewarded,\\
negligence is ever scorned.


\verseref{31}
The renunciate who delights in vigilance\\
and shuns heedlessness\\
advances like a grass fire,\\
consuming obstructions great and small.


\verseref{32}
The renunciate who delights in vigilance\\
and shuns heedlessness\\
is protected from regression:\\
such a one approaches liberation.

