
\chapter{The Thousands}


\verseref{100}
A single word of truth\\
which calms the mind\\
is better to hear than a thousand\\
irrelevant words.


\verseref{101}
A single verse of truth\\
which calms the mind\\
is better to hear than a thousand\\
irrelevant verses.


\verseref{102}
Reciting a single verse of truth\\
which calms the mind\\
is better than reciting\\
a hundred meaningless verses.


\verseref{103}
One might defeat alone in battle\\
a thousand thousand men,\\
but one who gains self-mastery\\
is by far the greater hero.


\verseref{104-105}
Self-mastery is the supreme victory --\\
much more to be valued\\
than winning control over others.\\
It is a victory\\
that no other being whatsoever\\
can distort or take away.


\verseref{106}
To honour, even for a single moment,\\
one who has attained self-mastery\\
is of greater merit\\
than making countless offerings\\
to those who are unworthy.


\verseref{107}
To honour, even for a single moment,\\
one who has attained self-mastery\\
is of greater benefit than a hundred years\\
spent habitually performing ceremonies\\
based on wrong understanding.

\verseref{108}
Even if one should spend\\
an entire year\\
making gestures of offering\\
for the sake of self-gain,\\
all this is not worth one quarter the value\\
of offering veneration to a noble being.


\verseref{109}
For one who always honours and respects\\
those of greater years\\
there are four blessings:\\
long life, beauty, happiness and strength.


\verseref{110}
A single day lived\\
with conscious intention and virtue\\
is of greater value than a hundred years\\
lived devoid of discipline and right restraint.


\verseref{111}
A single day lived\\
with conscious intention and wisdom\\
is of greater value than a hundred years\\
lived devoid of discipline and wisdom.


\verseref{112}
A single day lived\\
with conscious intention and profound effort\\
is of greater value than a hundred years\\
lived in lazy passivity.


\verseref{113}
A single day lived in awareness\\
of the transient nature of life\\
is of greater value than a hundred years\\
lived unaware of birth and death.


\verseref{114}
A single day lived awake to the undying state\\
is of greater value than a hundred years\\
lived without recognizing deathlessness.

\verseref{115}
A single day lived awake to profound truth\\
is of greater value than a hundred years\\
lived unawakened.

