
\chapter{The Awakened One}


\verseref{90}
There is no tension\\
for those who have completed their journey\\
and have become free\\
from the distress of bondage.


\verseref{91}
Alert to the needs of the journey,\\
those on the path of awareness,\\
like swans, glide on,\\
leaving behind their former resting places.


\verseref{92}
Just like birds that leave no tracks in the air,\\
there are those whose minds do not cling\\
to temptations that are offered to them.\\
Their focus is the signless state of liberation,\\
which to others is indiscernible.


\verseref{93}
There are those\\
who are free from all obstructions;\\
they do not worry about food,\\
their focus is the signless state of liberation.\\
Like birds flying through the air,\\
trackless they move on their way.


\verseref{94}
Like horses well-trained by their owners\\
are those who have rightly tamed their senses;\\
having freed their hearts\\
from pride and pollution\\
they generate all-pervading delight.


\verseref{95}
There are those who discover\\
they can completely abandon\\
confused reactions\\
and become patient as the earth;\\
unmoved by anger,\\
unshaken as a pillar,\\
unperturbed as a clear and quiet pool.


\verseref{96}
Those who arrive\\
at the state of perfect freedom\\
through right understanding\\
are unperturbed\\
in body, speech or mind.\\
They remain unshaken\\
by life's vicissitudes.


\verseref{97}
Those who know the uncreated,\\
who are free and stilled,\\
who have discarded all craving,\\
are the most worthy beings.


\verseref{98}
Whether in a forest,\\
a town or open country,\\
delightful is the dwelling place\\
of one now fully free.


\verseref{99}
Beings free from addiction to sensual pleasures\\
know a unique form of delight.\\
They seek quiet in forest retreats\\
which worldly beings would avoid.

