
\chapter{The Buddha}


\verseref{179}
The Buddha's perfection is complete;\\
there is no more work to be done.\\
No measure is there for his wisdom;\\
no limits are there to be found.\\
In what way could he be distracted\\
from truth?


\verseref{180}
The Buddha's perfection is complete;\\
in him there is no craving\\
that could drag him down.\\
No measure is there for his wisdom;\\
no limits are there to be found.\\
In what way could he be distracted\\
from truth?


\verseref{181}
Celestial beings treasure\\
the Awakened Ones\\
who have fully seen the Way,\\
who are devoted to meditation\\
and delight in the peace\\
of renunciation.


\verseref{182}
It is not easy to be born as a human being\\
and to live this mortal life.\\
It is not easy to have the opportunity\\
to hear Dhamma\\
and rare for a Buddha to arise.


\verseref{183}
Refrain from doing evil,\\
cultivate that which is good,\\
purify the heart.\\
This is the Way of the Awakened Ones.


\verseref{184}
A renunciate does not\\
oppress anyone.\\
Patient endurance\\
is the ultimate asceticism.\\
Profound liberation,\\
say the Buddhas,\\
is the supreme goal.


\verseref{185}
Not insulting, not harming,\\
cultivating restraint,\\
with respect for the training,\\
modesty in eating and contentment\\
with one's dwelling place,\\
devotion to mindful intent:\\
this is the teaching of the Buddha.


\verseref{186-187}
Not in great wealth\\
is there contentment,\\
nor in sensual pleasure,\\
gross or refined.\\
But in the extinction of craving\\
is joy to be found\\
by a disciple of the Buddha.


\verseref{188-189}
To many places beings withdraw\\
to escape from fear:\\
to mountains, forests,\\
parklands and gardens;\\
sacred places as well.\\
But none of these places\\
offer true refuge,\\
none of them can free us from fear.


\verseref{190-191}
One who finds refuge in the Buddha\\
in the Dhamma and in the Sangha\\
sees with penetrating insight:\\
suffering, its cause, its release\\
and the Way leading to true freedom.


\verseref{192}
The Buddha, the Dhamma, the Sangha:\\
these are the true refuge;\\
these are supreme;\\
these lead to Liberation.


\verseref{193}
It is hard to find\\
a being of great wisdom;\\
rare are the places\\
in which they are born.\\
Those who surround them\\
when they appear\\
know good fortune indeed.


\verseref{194}
Blessed is the arising of a Buddha;\\
blessed is the revealing of the Dhamma;\\
blessed is the concord of the Sangha;\\
delightful is harmonious communion.


\verseref{195-196}
Immeasurable is the benefit\\
obtained from honouring those\\
who are pure and beyond fear.\\
Beings who have found freedom\\
from sorrow and grieving\\
are worthy of honour.

