
\chapter{The Self}


\verseref{157}
If we hold ourselves dear,\\
then we maintain careful self-regard\\
both day and night.


\verseref{158}
It is wise\\
to set ourselves right\\
before instructing others.


\verseref{159}
One's own self is the hardest to discipline.\\
You should act as you teach:\\
tame yourself\\
before trying to tame others.


\verseref{160}
Truly it is ourselves\\
that we depend upon;\\
how could we really\\
depend upon another?\\
When we reach the state\\
of self-reliance\\
we find a rare refuge.


\verseref{161}
As a diamond can cut through\\
the stone that once housed it,\\
so your own evil can grind you down.


\verseref{162}
Committed evil-doers\\
behave towards themselves\\
like their own worst enemies.\\
They are like creepers\\
that strangle the trees\\
which support them.


\verseref{163}
It is easy to do that which is\\
of no real benefit to oneself,\\
but it is difficult indeed to do that\\
which is truly beneficial and good.


\verseref{164}
Like the bamboo\\
which destroys itself as it bears fruit,\\
so fools harm themselves\\
by holding to wrong views\\
and deriding those worthy ones\\
who live in harmony with the Way.


\verseref{165}
By ourselves we do evil\\
and by ourselves we are made impure.\\
By ourselves we avoid evil\\
and by ourselves we are made pure.\\
The great matter of purity\\
is our own affair.\\
No other can be responsible.


\verseref{166}
Knowing the Way for oneself,\\
walk it thoroughly.\\
Do not allow the needs of others,\\
however demanding,\\
to bring about distraction.

