
\chapter{Old Age}


\verseref{146}
Why is there laughter?\\
Why is there joy\\
when the world is on fire?\\
Since you are clouded in darkness\\
should you not seek the light?


\verseref{147}
Gaze upon this once-decorated body --\\
it used to attract attention\\
but now it is only festering flesh,\\
a putrid thing.\\
It is neither sure nor substantial.


\verseref{148}
This body wears out with age;\\
it becomes a host to disease\\
- vulnerable, fragile,\\
a decrepit, disintegrating mass,\\
which eventually ends in death.


\verseref{149}
What pleasure does life hold\\
once one has seen\\
old bleached bones\\
discarded and scattered round?


\verseref{150}
The physical body consists of bones\\
covered with flesh and blood.\\
Stored up inside it\\
are decay and death, pride and malice.


\verseref{151}
Passed down by the wise\\
is the knowledge that\\
though what is externally impressive\\
loses its splendour,\\
and though our bodies will decay,\\
the truth itself outlasts all degeneration.

\verseref{152}
While ageing fools put on weight\\
like oxen in their stalls,\\
their minds remain small.


\verseref{153-154}
For many lives I have wandered\\
looking for, but not finding,\\
the house-builder\\
who caused my suffering.\\
But now you are seen and\\
you shall build no more.\\
Your rafters are dislodged and\\
the ridge-pole is broken.\\
All craving is ended;\\
my heart is as one with the unmade.


\verseref{155}
Those who, while still young,\\
neither choose a life of renunciation\\
nor earn a good living,\\
end up like dejected old herons\\
beside a pond without fish.


\verseref{156}
Those who, while still young,\\
neither choose a life of renunciation\\
nor earn a good living,\\
will end up bemoaning the past,\\
falling like spent arrows\\
that have missed their mark.

